A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-08-23 paper/project.tex
Edits to some energy plots for 1D slit and metastability discussion

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
\needsworklater{ We develop a classical density functional theory for
  water by combining the fundamental-measure theory functional--which
  describes hard sphere fluids--with attractive interactions based on the 
  Statistical Associating Fluid Theory (SAFT).  Using
  this approximation, we are able to describe various properties of
  water near liquid-vapor and liquid-solid interfaces including water confined 
  to a slit and two hydrophobic rods immersed in water.  This functional will be the 
  foundation for a efficient method for future work
  predicting the interactions of chemicals in aqueous solution on an
  atomic level.}

@@

The simple yet complex water molecule is one of the most important 
substances for life and industry. The structure and geometry of the 
water molecule leads to a strong polarity and complex intermolecular
forces. The hydrogen bonds formed by water are very strong, creating
tetragonal clusters (CITE?) that lead to many of water's unusual properties,
such as the comparatively high melting, boiling, and critical temperatures.
The large dielectric constant (D~80) (CITE) is partially responsible for
water as a universal solvent, along with the ease of availability and use.

@@

(Previous/popular models of water summary/review??)

@@

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/entropy}
% \end{center}
% \caption{The free energy, internal energy, and temperature*entropy versus density at room temperature.  }
% \label{fig:energy-room-temp}
% \end{figure}
% 
% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/entropy-at-690K}
% \end{center}
% \caption{The free energy, internal energy, and temperature*entropy versus density near the critical temperature.  }
% \label{fig:energy-near-critical-temp}
% \end{figure}

@@

\subsection{Determining the empirical parameters (?)}\label{sec:empirical}

@@

The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the \emph{exact}
density-functionaly theory \emph{would} include them), with the result
that the density profile will be smoothed out by CW effects.

\textcolor{red}{I found a nice paper on the hard-slit
  problem... actually a couple of papers that address
  it~\cite{forsman1996computer, lum1999hydrophobicity}}

@@

We choose a chemical potential corresponding to a pressure of 1 atm and we apply 
a constraining potential where the density is equal to the saturated 
liquid density inside the slit. Outside the slit, we choose a small non-zero
density for numerical reasons. The slits studied range in size from 0.1 nm to 
8.0 nm. We minimize the energy and calculate density, energy, and $X$ on a lattice
with a spacing of approximately 0.01 nm (0.2 bohrs) between lattice points.

The density profile for a slit of size 4.0 nm is shown in Figure 
\ref{fig:density-1D}. Near the interface, there are oscillations around the 
liquid density which decrease near the center of the slit. We expect the
center of the slit to act approximately like bulk water, and the absence 
of oscillations supports this. For larger slits (6.0 nm or more) the
majority of the water molecules in the slit act like the bulk. We also 
determine the smallest slit that can contain liquid water, approximately 
0.6 nm (\textcolor{red}{Check this again}. For smaller slits, the initial 
liquid water inside will vaporize.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-1D}
\end{center}
\caption{ Energy densities for a slab of water with molecules 
constrained to move in one dimension. \textcolor{red}{FIXME}}
\label{fig:energy-1D}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/cavitysize-vs-energy-1D}
\end{center}
\caption{ Free energy versus slit size for a slab of water with molecules 
constrained to move in one dimension. \textcolor{red}{Not sure if this needs
to have a bulk energy amount removed to be accurate.}}
\label{fig:cavitysize-vs-energy-1D}
\end{figure}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
\needsworklater{ We develop a classical density functional theory for
  water by combining the fundamental-measure theory functional--which
  describes hard sphere fluids--with attractive interactions based on the 
  Statistical Associating Fluid Theory (SAFT).  Using
  this approximation, we are able to describe various properties of
  water near liquid-vapor and liquid-solid interfaces including water confined 
  to a slit and hydrophobic rods immersed in water.  This functional will be
  the 
  foundation for a efficient method for future work
  predicting the interactions of chemicals in aqueous solution on an
  atomic level.}

@@

The simple yet complex water molecule is one of the most important 
substances for life and industry. The structure and geometry of the 
water molecule leads to a strong polarity and complex intermolecular
forces. The hydrogen bonds formed by water are very strong, creating
tetragonal clusters (CITE?) that lead to many of water's unusual properties,
such as the comparatively high melting, boiling, and critical temperatures.
The large dielectric constant (D$\sim$80) (CITE) is partially responsible for
water as a universal solvent, along with the ease of availability and use.

@@

(Types and models of water summary/review)

@@

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/entropy}
% \end{center}
% \caption{The free energy, internal energy, and temperature*entropy 
% versus density at room temperature.  }
% \label{fig:energy-room-temp}
% \end{figure}

@@

\subsection{Determining the empirical parameters}\label{sec:empirical}

@@

The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the \emph{exact}
density-functional theory \emph{would} include them), with the result
that the density profile will be smoothed out by CW effects.

@@

We choose a chemical potential corresponding to a pressure of 1 atm and set the
temperature to a typical room temperature of 298 K. We apply 
a constraining potential outside the slit, where we choose a small non-zero
density for numerical reasons. The slits studied range in size from 0.1 nm to 
8.0 nm. We minimize the energy and calculate density, energy, and $X$ on a
lattice
with a spacing of approximately 0.01 nm (0.2 bohrs) between lattice points.

The density profile for a slit of size 4.0 nm is shown in Figure
\ref{fig:density-1D}. Near the interface, there are oscillations around the
liquid density which decrease near the center of the slit. We expect the center
of the slit to act approximately like bulk water, and the absence of
oscillations supports this. For larger slits (6.0 nm or more) the majority of
the water molecules in the slit act like the bulk. Based on our computations, we
also determine the smallest slit that can contain liquid water. This result is
approximately 0.6 nm. For smaller slits, the initial liquid water inside will
vaporize. The results for larger liquid-filled slits are only metastable,
however,as a vapor-filled slit is thermodynamically favorable. Based on our
calculations for a pressure of 1 atm, the work needed to pull apart
the two hard walls comprising the slit is one to two orders of magnitude less
thanthe free energy of the liquid-filled slit. For the range of slit sizes we
studied (0.1 to 8.0 nm), all the computational results of liquid-filled are
metastable. Lum, Chandler and Weeks \cite{lum1999hydrophobicity} discuss the
stability of water between large hydrophobic plates and the critical separation
necessary for stability. For water at room temperature and atmospheric pressure,
plates with a separation smallerthan about 100 nm will not contain stable liquid
water. They go on to say that confined water has a region of metastability down
to separations of about 5 nm. This is comparable to the slit separation where
our calculations show water inside the slit acting like the bulk liquid, which
we know to be a metastable solution. Forsman et al.\cite{forsman1996computer}
also study water between two hard walls at constant chemical potential. They 
also conclude that the liquid state is likely metastable, and there is a free
energy barrier to reach the stable vapor phase.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-1D}
\end{center}
\caption{ Energy densities for a slab of water with molecules 
constrained to move in one dimension. The free energy, internal energy
and entropic energy densities are shown for a slit size of 4.0 nm.}
\label{fig:energy-1D}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/cavitysize-vs-energy-1D}
\end{center}
\caption{ Energy per length versus slit size for a slab of water with molecules 
constrained to move in one dimension. The last few data points are likely inaccurate
due to the slit separation being very close to the size of the system. \textcolor{red}{
Not sure if this needs to have a bulk energy amount removed to be accurate.}}
\label{fig:cavitysize-vs-energy-1D}
\end{figure}
